The Role of Ice Compositions and Morphology for Snowlines and the C/N/O Ratios in Active Disks

### 2016-03-09 tex/draft2/ms_edited.tex
update

\author{Ana-Maria A. Piso\altaffilmark{1}, Karin I. \"Oberg\altaffilmark{1}, Jamila Pegues\altaffilmark{2}}
\altaffiltext{1}{Harvard-Smithsonian Center for Astrophysics, 60 Garden Street, Cambridge, MA 02138}
\altaffiltext{2}{Department of Astrophysical Sciences, Princeton University}

@@

Both in Solar system comets and in protoplanetary disks carbon and oxygen are primarily contained in H$_2$O, CO$_2$ and CO (e.g., \citealt{lodders03}, \citealt{mumma11}, \citealt{oberg11}, \citealt{boogert15}). However, some fraction of the carbon abundance may also be carried by CH$_4$ (e.g., \citealt{oberg08}), which may change the C/O ratio in gas and in dust throughout the disk. In the case of nitrogen, chemical models of the protostellar nebula (e.g., \citealt{owen01}) and of protoplanetary disks (e.g., \citealt{rodgers02}) suggest that N$_2$ was the dominant form of nitrogen, and that giant planets have accreted their nitrogen content primarily as N$_2$ \citep{mousis14}. Observations of Solar system bodies such as Titan and Pluto show that N$_2$ is prevalent in their atmospheres (\citealt{cruikshank93}, \citealt{owen93}).  Moreover, the Rosetta spacecraft has recently made the first direct measurement of the N$_2$ abundance in comet 67P/Churyumov-Gerasimenko \citep{rubin15}. In addition to N$_2$, a fraction of the nitrogen abundances may be also carried by NH$_3$ (\citealt{bottinelli10}, \citealt{mumma11}). Because of the high volatility of N$_2$, the gas phase nitrogen-to-oxygen (N/O) ratio in the outer disk may be even more enhanced than the C/O ratio compared to its average value in the disk. Giant planets that form at wide separations should thus have an excess of nitrogen in their atmospheres, which could be used to trace their formation origin.

Moreover, snowline locations strongly depend on the ice grain morphology. Very volatile species, such as CO and N$_2$, present binding energies, and therefore snowline locations, that are sensitive to the details of the morphology of the icy grain mantles. Laboratory experiments (\citealt{collings03}, \citealt{oberg05}, \citealt{bisschop06}, \citealt{fayolle16}) have shown that CO and N$_2$ have significantly different binding energies depending on whether they are pure or water dominated ices. This implies that ices in different environments will sublimate at different radii, which will substantially change the disk regions where these volatiles are present in gaseous or solid form (see Section \ref{sec:static}).

@@

In this work, we expand the coupled drift-desorption model developed in \citet{piso15b} (hereafter Paper I) by considering additional volatile molecules and abundances, ice morphology, as well as nitrogen-to-oxygen (N/O) ratios.

This paper is organized as follows. In Section \ref{sec:review}, we review the drift-desorption model developed in Paper I. We discuss the effect of different abundances of the main carbon, oxygen and nitrogen carriers, grain morphology and disk dynamics on snowline locations and the C/N/O ratios in Section \ref{sec:results}. We address the implications of our results in Section \ref{sec:discussion} and summarize our findings in Section \ref{sec:summary}.

@@

As we note in Section \ref{sec:intro}, the disk volatile composition and the ice morphology determine the location of several important snowlines. In this work we focus on the primary carbon, oxygen and nitrogen carriers, i.e. H$_2$O, CO$_2$, CO, N$_2$, and to a lesser extent, CH$_4$ and NH$_3$. To quantify the effect of the presence of some carbon in the form of CH$_4$ and of some nitrogen in the form of NH$_3$, we use measured median CH$_4$ and NH$_3$ abundances in protostellar cores from the \textit{Spitzer} c2d Legacy ice survey \citep{evans03}.

We determine the location of the H$_2$O, CO$_2$, CO, CH$_4$, N$_2$ and NH$_3$ snowlines in our static disk by balancing desorption with readsorption, following \citet{hollenbach09}. The binding energies of H$_2$O, CO$_2$, CO, CH$_4$, N$_2$ and NH$_3$ as pure ices are 5800 K, 2000 K, 834 K, 1300 K, 767 K and 2965 K, respectively (\citealt{fraser01}, \citealt{collings04}, \citealt{fayolle16}, \citealt{garrod06}, \citealt{martin14}). For CO and N$_2$ as water dominated ices, the binding energies are 1388 K and 1266 K, respectively \citep{fayolle16}. Figure \ref{fig:CNOstatic} shows the resulting snowline locations, with CO and N$_2$ pure ices (top panel) and water dominated ices (bottom panel). The ordinate displays the total carbon, oxygen and nitrogen abundance in solids as a function of the hydrogen total abundance.  As expected, the total grain abundance increases with semimajor axis, as more and more species sublimate. Freeze-out at the CO$_2$ and CO snowlines pulls more heavy elements into the grains than in the case of the H$_2$O snowline. More importantly, the CO and N$_2$ snowlines move several tens of AU inward if the ices are water dominated rather than pure. This changes the chemical abundances both in gas and dust throughout the disk, directly affecting the compositions of nascent giant planets forming in situ.

@@

In this section we determine the C/O and N/O ratios in gas and dust throughout our static disk, and the extent to which they are affected by the presence of CH$_4$ and NH$_3$. Since here we are primarily interested in the magnitude of the C/N/O ratios rather than snowline locations (which have been discussed in Section \ref{sec:snowlines}), we assume for simplicity that CO and N$_2$ are pure ices.

We explore the parameter space of possible CH$_4$ abundances by assuming three different scenarios: (1) no CH$_4$, (2) the median CH$_4$ observed abundance (hereafter CH$_4$-mid), and (3) the maximum CH$_4$ observed abundance (hereafter CH$_4$-max). Thus $n_{\rm CH_4-mid}=0.0555 \times n_{\rm H_2O}$ \citep{oberg11a} and $n_{\rm CH_4-max}=0.13 \times n_{\rm H_2O}$ \citep{oberg08}, where $n_{\rm H_2O}$ is the total H$_2$O abundance. Similarly to Paper I, we use the H$_2$O, CO$_2$ and CO abundances of \citet{oberg11}. Since the abundance of carbon grains is uncertain, we assume that all the carbon that is not in the form of CH$_4$, CO and CO$_2$ is in carbon grains, so that we reproduce the Solar C/O ratio (gas+dust) of 0.54.

@@

We assume that the main nitrogen-bearing species are N$_2$ and NH$_3$, since other volatiles that contain nitrogen have significantly lower abundances in comparison (e.g., \citealt{mumma11}). We use the measured total nitrogen abundance in the Solar system, $n_{\rm N}=8 \times 10^{-5} n_{\rm H}$ \citep{lodders03}, where $n_{\rm H}$ is the hydrogen abundance in the disk midplane. Similarly to the case of CH$_4$, we explore the parameter space of possible NH$_3$ abundances using data from the Spitzer c2d Legacy ice survey, as follows: (1) no NH$_3$, (2) the median NH$_3$ observed abundance $n_{\rm NH_3-mid}=0.055 \times n_{\rm H_2O}$ \citep{oberg11a}, and (3) the maximum observed NH$_3$ abundance $n_{\rm NH_3-max}=0.15 \times n_{\rm H_2O}$ \citep{bottinelli10}. In each case, the N$_2$ abundance then simply follows as $n_{\rm N_2}=(n_{\rm N}-n_{\rm NH_3})/2$.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{N_O_ratio.pdf}
%\vspace{-0.5in}
\caption{The N/O ratio in gas (solid lines) and dust (dashed lines) as a function of semimajor axis in a static disk, assuming no nitrogen is present in the form of NH$_3$ (top panel), the median observed NH$_3$ abundance is assumed (middle panel), and the maximum observed NH$_3$ abundance is assumed (bottom panel). The N/O estimates are performed assuming that the CO and N$_2$ ices are in pure form. The vertical dashed lines mark the snowline locations of the main C,O and N carriers. The horizontal dashed lines represent the average N/O value in the disk. The gas-phase N/O ratio is highly enhanced by a factor of three between the CO$_2$ and CO snowlines compared to its average value. The arrows mark a highly elevated N/O ratio in gas between the CO and N$_2$ snowlines due to the depletion of oxygen gas in this region. The presence of NH$_3$ moderately decreases the N/O ratio in gas between the NH$_3$ and CO$_2$ snowlines.} 
\label{fig:Nstatic}
\end{figure}

@@

Figure \ref{fig:NO_ratio} shows the H$_2$O, CO$_2$, CO and N$_2$ snowline locations in a viscous disk for the same range of initial particle sizes as in Figure \ref{fig:CO_ratio}, and with the CO and N$_2$ snowlines calculated assuming different grain morphologies as explained above, as well as estimates for the N/O ratio throughout the disk. For simplicity, we assume that all nitrogen is bound in N$_2$. This choice is justified since the presence of some NH$_3$ only moderately changes the N/O ratio (see Figure \ref{fig:Nstatic}), and since we are primarily interested in the N$_2$ snowline locations rather than exact values for the N/O ratio. The innermost N$_2$ snowlines in the viscous disk, created by particles with $s \sim 7$ m for our fiducial model, are located at $r_{\rm N_2, pure} \approx 42$ AU for N$_2$ as pure ice and at $r_{\rm N_2, water} \approx 11$ AU for N$_2$ in water dominated ices. Thus for each case (pure versus water dominated ices), the N$_2$ snowline moves inward by up to 50\% due to disk dynamics. By taking into account both ice morphology and disk dynamics, the full range of potential N$_2$ snowlines span $\sim$11 to $\sim$79 AU, which is a factor of $\sim$7 difference. Similarly to the case for CO, the N$_2$ snowline may be close to 10 AU for certain disk models.

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{N_O_water_ice.pdf}
%\vspace{-0.5in}
\caption{N/O ratio estimates in gas (solid lines) and dust (dashed lines) as function of semimajor axis in a viscous disk, for CO and N$_2$ as pure ices (top panel) or as water dominated ices (bottom panel). The H$_2$O, CO$_2$ and N$_2$ snowlines are shown for particles with initial sizes $\sim0.001$ cm $\lesssim s \lesssim$ 7 m as indicated by the color bar. The CO snowline (inner vertical dashed line in both panels) is only shown for the 7 m particles. The N/O ratio in a static disk (black lines) is shown for comparison. The downward arrows show that the N/O ratio in gas will decrease inside the H$_2$O and CO$_2$ snowlines in the viscous disk, as the relative fluxes of the desorbed icy
particles and the overall nebular gas will cause an excess of oxygen gas inside these snowlines (see Paper I for details). Radial drift and gas accretion move the N$_2$ snowline inward by up to $\sim$50\% compared to a static disk. The presence of N$_2$ in a water ice environment rather than as pure ice moves the N$_2$ snowline significantly inward by $\sim$70\%. Taken together, disk dynamics and ice morphology move the CO snowline inward by a factor of $\sim$7. The results of highly enhanced gas-phase N/O ratios between the CO$_2$ and CO snowlines compared to its average value, and of highly elevated N/O ratios in gas between the CO and N$_2$ snowlines (see Figure \ref{fig:Nstatic}), are preserved.}  
\label{fig:NO_ratio}
\end{figure}

@@

%\emgr{Discuss how entrapment of volatiles by H2O affects volatile abundances and C/O ratios. Re-emphasize the fact that the C/O and N/O ratios are upper estimates, and that CH4 and NH3 might matter in a viscous disk. State that we plan to address this in a future paper. More TBD.}
This study shows that the gas-phase N/O ratio in protoplanetary disks is considerably enhanced throughout most of the disk midplane compared to its average value. As demonstrated in Figure \ref{fig:enhance}, the gaseous N/O ratio is enhanced by a factor of two beyond the H$_2$O snowline, by more than a factor of three between the CO$_2$ and CO snowlines, and by several orders of magnitude between the CO and N$_2$ snowlines. Thus constraining the N/O ratio in a giant planet atmosphere could be used to trace its formation origins. Theoretical models of the magnitude and role of N/O (and N/C) ratios in exoplanet atmospheres are therefore key in order to use these ratios as probes for a planet's formation location. Measurements of the N/O ratio in planet atmospheres are essential to further constrain the theoretical expectations --- while nitrogen-bearing molecules have not been targeted so far due to lack of instrument sensitivity, such observations and detections are likely in the near future with the advent of JWST (e.g., NH$_3$, \citealt{greene16}).

@@

The N/O ratio enhancement is larger than that of the gas phase C/O ratio throughout most of the disk. Thus measurements of an enhanced C/O ratio in an exoplanet atmosphere could be corroborated (disproved) by measurements of enhanced (non-enhanced) N/O ratios. Moreover, Figure \ref{fig:enhance} shows that giant planets that have formed in situ between the H$_2$O and CO snowlines are expected to present elevated both C/O and N/O ratios in their atmospheres, whereas planets between the CO and N$_2$ snowlines will have a highly enhanced N/O ratio in their atmospheres, but not C/O.

Due to disk dynamics and ice morphology, the disk regions with highly elevated gas phase N/O and C/O ratios are uncertain and may span tens of AU. In particular, one can imagine a scenario where CO is in a water binding environment and N$_2$ is not. This could be attributed to the fact that H$_2$O may bind preferentially to CO than N$_2$, since both H$_2$O and CO are polar molecules while N$_2$ is not. It is also possible for N$_2$ ices to form later than CO (e.g., \citealt{pagani12}), and thus be deposited on the outer layers of the icy mantles which are typically water poor (e.g., \citealt{garrod11}). Detections of CO and N$_2$ snowlines in the same disk could potentially verify this scenario, since CO and N$_2$ desorb at substantially different temperature, and therefore locations, if the former is water dominated and the latter is in a pure ice environment. Changes in stellar luminosity (e.g., \citealt{kennedy06}) and gas mass accretion rate (e.g., \citealt{chambers09}) may introduce additional uncertainties in the snowline locations, and thus the C/N/O ratios (see also Paper I). The impact of the ice environment on the snowline location is much smaller in the case of CO$_2$ and NH$_3$, as their binding energies and behavior are closer to that of H$_2$O. No detailed measurements for the CH$_4$ binding energy in a water environment exist so far, but due to its low desorption temperature a similar behavior to that of CO and N$_2$ would be expected. While the presence of some carbon in the form of CH$_4$ only modestly affects our results, CH$_4$ may become important in disks where a large fraction of the CO abundance has been converted into hydrocarbons (e.g., \citealt{du15}).  %It is thus crucial to build a quantitative framework to explore the range of snowline locations and thus gaseous C/N/O ratios across disks with different dynamics and grain morphologies, as such models are currently lacking. %The theoretical expectations for N/O ratios in disks and giant planets may be further constrained by future JWST observations of exoplanet atmospheric spectra and detections of nitrogen bearing species (e.g., NH$_3$, \citealt{greene16}).

Our results could also be applied to the Solar system. Recent measurements of nitrogen abundance in comet 67P/Churyumov-Gerasimenko found a N$_2$/CO ratio $\sim 10^{-3}$ \citep{rubin15}. A low N$_2$/CO ratio is consistent with comets having formed inside the N$_2$ snowline where solid N$_2$ is depleted. Theoretical models suggest that Jupiter-family comets, such as 67P, originate from the Kuiper belt \citep{duncan97} (but see \citealt{rubin15} for alternative formation scenarios for 67P). It is thus possible, in principle, to use measurements of the N$_2$ abundance in Jupiter-family comets to determine where the N$_2$ snowline was located in our Solar system. However due to the uncertainty in the calculated location of the N$_2$ snowline (see Section \ref{sec:dynamic}), more detailed modeling is needed.

Finally, detections of snowlines in disks that are reasonably well characterized could help further constrain the dynamics and ice morphology in the disks.

@@

\item The presence of some carbon in the form of CH$_4$ only modestly affects our results for the gas phase C/O ratio, increasing it by up to 8\% between the CO$_2$ and CH$_4$ snowlines. The presence of some nitrogen in the form of NH$_3$ moderately affects our results and may decrease the gaseous N/O ratio by up to 18\% between the NH$_3$ and CO$_2$ snowlines. In both cases, large gas phase C/O and N/O ratios in the outer disk are preserved.

>>>>

\author{Ana-Maria A. Piso\altaffilmark{1}, Jamila Pegues\altaffilmark{2}, Karin I. \"Oberg\altaffilmark{1}}
\altaffiltext{1}{Harvard-Smithsonian Center for Astrophysics, 60 Garden Street, Cambridge, MA 02138}
\altaffiltext{2}{Department of Astrophysical Sciences, Princeton University}

@@

Both in Solar system comets and in protoplanetary disks, volatile carbon and oxygen are primarily contained in H$_2$O, CO$_2$ and CO (e.g., \citealt{lodders03}, \citealt{mumma11}, \citealt{oberg11}, \citealt{boogert15}). However, some fraction of carbon may also be carried by CH$_4$ (e.g., \citealt{oberg08}), which may change the C/O ratio in gas and in dust throughout the disk. In the case of nitrogen, chemical models of the protostellar nebula (e.g., \citealt{owen01}) and of protoplanetary disks (e.g., \citealt{rodgers02}) suggest that N$_2$ was the dominant form of nitrogen, and that giant planets have accreted their nitrogen content primarily as N$_2$ \citep{mousis14}. Observations of Solar system bodies such as Titan and Pluto show that N$_2$ is prevalent in their atmospheres (\citealt{cruikshank93}, \citealt{owen93}).  Moreover, the Rosetta spacecraft has recently made the first direct measurement of the N$_2$ abundance in comet 67P/Churyumov-Gerasimenko \citep{rubin15}. In addition to N$_2$, a fraction of the nitrogen abundances may be also carried by NH$_3$ (\citealt{bottinelli10}, \citealt{mumma11}). Because of the high volatility of N$_2$, the gas phase nitrogen-to-oxygen (N/O) ratio in the outer disk may be even more enhanced than the C/O ratio compared to its average value in the disk. Giant planets that form at wide separations should thus have an excess of nitrogen in their atmospheres, which could be used to trace their formation origin.

The snowline locations of the main carbon, oxygen and nitrogen carriers strongly depend on the ice grain morphology. Very volatile species, such as CO and N$_2$, present binding energies, and therefore snowline locations, that are sensitive to the details of the morphology of the icy grain mantles. Laboratory experiments (\citealt{collings03}, \citealt{oberg05}, \citealt{bisschop06}, \citealt{fayolle16}) have shown that CO and N$_2$ have significantly different binding energies depending on whether they are pure or water dominated ices. This implies that ices in different environments will sublimate at different radii, which will substantially change the disk regions where these volatiles are present in gaseous or solid form (see Section \ref{sec:static}).

@@

In this work, we expand the coupled drift-desorption model developed in \citet{piso15b} (hereafter Paper I) by considering additional volatile molecules and abundances, ice morphology, as well as nitrogen-to-oxygen (N/O) ratios. This paper is organized as follows. In Section \ref{sec:review}, we review the drift-desorption model developed in Paper I. We discuss the effect of different abundances of the main carbon, oxygen and nitrogen carriers, grain morphology and disk dynamics on snowline locations and the C/N/O ratios in Section \ref{sec:results}. We address the implications of our results in Section \ref{sec:discussion} and summarize our findings in Section \ref{sec:summary}.

@@

As we note in Section \ref{sec:intro}, the disk volatile composition and the ice morphology determine the location of important snowlines. In this work we focus on the primary carbon, oxygen and nitrogen carriers, i.e. H$_2$O, CO$_2$, CO, N$_2$, and to a lesser extent, CH$_4$ and NH$_3$. %To quantify the effect of the presence of some carbon in the form of CH$_4$ and of some nitrogen in the form of NH$_3$, we use measured median CH$_4$ and NH$_3$ abundances in protostellar cores from the \textit{Spitzer} c2d Legacy ice survey \citep{evans03}. 
Our standard model is based on the median ice abundances observed toward Solar-type protostars \citep{oberg11a}, which are $n_{\rm CO_2}=0.29 \times n_{\rm H_2O}$, $n_{\rm CO}=0.38 \times n_{\rm H_2O}$, $n_{\rm CH_4}=0.0555 \times n_{\rm H_2O}$ (hereafter CH$_4$-mid) and $n_{\rm NH_3}=0.055 \times n_{\rm H_2O}$ (hereafter NH$_3$-mid). Here $n_{\rm H_2O} \approx 10^{-4} \times n_{\rm H}$ is the total water abundance \citep{vandishoeck06}, with $n_{\rm H}$ the hydrogen abundance in the disk midplane. For CO, we also take into account that the observed CO ice only traces some of the CO reservoir due to its high volatility, and similarly to \citet{oberg11} and Paper I we set the total CO abundance to $0.9 \times 10^{-4} n_{\rm H}$. Finally, we assume that all nitrogen not found in NH$_3$ is in N$_2$ and assume a Solar nitrogen abundance, $n_{\rm N}=8 \times 10^{-5} n_{\rm H}$ \citep{lodders03}. In effect, this model assumes no chemical evolution between the protostellar and disk midplane stages. This is reasonable for material that accretes onto the disk at large radii \citep{visser09}, but may overestimate the contribution of the original volatiles to the total volatile budget in the innermost disk.

We determine the location of the H$_2$O, CO$_2$, CO, CH$_4$, N$_2$ and NH$_3$ snowlines in our static disk by balancing desorption with readsorption, following \citet{hollenbach09}. The binding energies of H$_2$O, CO$_2$, CO, CH$_4$, N$_2$ and NH$_3$ as pure ices are 5800 K, 2000 K, 834 K, 1300 K, 767 K and 2965 K, respectively (\citealt{fraser01}, \citealt{collings04}, \citealt{fayolle16}, \citealt{garrod06}, \citealt{martin14}). For CO and N$_2$ as water dominated ices, the binding energies are 1388 K and 1266 K, respectively \citep{fayolle16}. Figure \ref{fig:CNOstatic} shows the resulting snowline locations, with CO and N$_2$ pure ices (top panel) and water dominated ices (bottom panel). The ordinate displays the total carbon, oxygen and nitrogen abundance in solids as a function of the hydrogen total abundance.  As expected, the total grain abundance increases with semimajor axis, as more and more species freeze out. Freeze-out at the CO$_2$ and CO snowlines pulls more heavy elements into the grains than in the case of the H$_2$O snowline.  The CO and N$_2$ snowlines move several tens of AU inward if the ices are water dominated rather than pure. This changes the chemical abundances both in gas and dust throughout the disk, directly affecting the compositions of nascent giant planets forming in situ.

@@

In this section we determine the C/O and N/O ratios in gas and dust throughout our static disk, and to what extent they are affected by the presence of CH$_4$ and NH$_3$ over the full range of observed CH$_4$ and NH$_3$ abundances. Since here we are primarily interested in the magnitude of the C/N/O ratios rather than snowline locations (which have been discussed in Section \ref{sec:snowlines}), we assume for simplicity that CO and N$_2$ are pure ices.

We explore the parameter space of possible CH$_4$ abundances by assuming three different scenarios: (1) no CH$_4$, (2) CH$_4$-mid, and (3) the maximum CH$_4$ observed abundance (hereafter CH$_4$-max), $n_{\rm CH_4-max}=0.13 \times n_{\rm H_2O}$ \citep{oberg08}, where $n_{\rm H_2O}$ is the total H$_2$O abundance. %Similarly to Paper I, we use the H$_2$O, CO$_2$ and CO abundances of \citet{oberg11}. 
Since the abundance of carbon grains is uncertain, we assume that all the carbon that is not in the form of CH$_4$, CO and CO$_2$ is in carbon grains, so that we reproduce the Solar C/O ratio (gas+dust) of 0.54.

@@

We assume that the main nitrogen-bearing species are N$_2$ and NH$_3$, since other volatiles that contain nitrogen have significantly lower abundances in comparison (e.g., \citealt{mumma11}). Similarly to the case of CH$_4$, we explore the parameter space of possible NH$_3$ abundances using observations of protostellar cores, as follows: (1) no NH$_3$, (2) NH$_3$-mid, and (3) the maximum observed NH$_3$ abundance $n_{\rm NH_3-max}=0.15 \times n_{\rm H_2O}$ \citep{bottinelli10}. In each case, the N$_2$ abundance then simply follows as $n_{\rm N_2}=(n_{\rm N}-n_{\rm NH_3})/2$.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{N_O_ratio.pdf}
%\vspace{-0.5in}
\caption{The N/O ratio in gas (solid lines) and dust (dashed lines) as a function of semimajor axis in a static disk, assuming no nitrogen is present in the form of NH$_3$ (top panel), the median observed NH$_3$ abundance is assumed (middle panel), and the maximum observed NH$_3$ abundance is assumed (bottom panel). The N/O estimates are performed assuming that the CO and N$_2$ ices are in pure form. The vertical dashed lines mark the snowline locations of the main C,O and N carriers. The horizontal dashed lines represent the average N/O value in the disk. The gas-phase N/O ratio is enhanced by a factor of two between the H$_2$O and CO$_2$ snowlines compared to its average value, and by a factor of three between the CO$_2$ and CO snowlines. The arrows mark a highly elevated N/O ratio in gas between the CO and N$_2$ snowlines due to the depletion of oxygen gas in this region. The presence of NH$_3$ moderately decreases the N/O ratio in gas between the NH$_3$ and CO$_2$ snowlines.} 
\label{fig:Nstatic}
\end{figure}

@@

Figure \ref{fig:NO_ratio} shows the H$_2$O, CO$_2$, CO and N$_2$ snowline locations in a viscous disk for the same range of initial particle sizes as in Figure \ref{fig:CO_ratio}, and with the CO and N$_2$ snowlines calculated assuming different grain morphologies as explained above, as well as estimates for the N/O ratio throughout the disk. For simplicity, we assume that all nitrogen is the form of N$_2$. This choice is justified since the presence of some NH$_3$ only moderately changes the N/O ratio (see Figure \ref{fig:Nstatic}), and since we are primarily interested in the N$_2$ snowline locations rather than exact values for the N/O ratio. The innermost N$_2$ snowlines in the viscous disk, created by particles with $s \sim 7$ m for our fiducial model, are located at $r_{\rm N_2, pure} \approx 42$ AU for N$_2$ as pure ice and at $r_{\rm N_2, water} \approx 11$ AU for N$_2$ in water dominated ices. Thus for each case (pure versus water dominated ices), the N$_2$ snowline moves inward by up to 50\% due to disk dynamics. By taking into account both ice morphology and disk dynamics, the full range of potential N$_2$ snowlines span $\sim$11 to $\sim$79 AU, which is a factor of $\sim$7 difference. Similarly to the case for CO, the N$_2$ snowline may be close to 10 AU for certain disk models.

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{N_O_water_ice_many.pdf}
%\vspace{-0.5in}
\caption{N/O ratio estimates in gas (solid lines) and dust (dashed lines) as function of semimajor axis in a viscous disk, for CO and N$_2$ as pure ices (top panel) or as water dominated ices (bottom panel). The H$_2$O, CO$_2$, CO and N$_2$ snowlines are shown for particles with initial sizes $\sim0.001$ cm $\lesssim s \lesssim$ 7 m as indicated by the color bar. The N/O ratio in a static disk (black lines) is shown for comparison. The downward arrows show that the N/O ratio in gas will decrease inside the H$_2$O and CO$_2$ snowlines in the viscous disk, as the relative fluxes of the desorbed icy
particles and the overall nebular gas will cause an excess of oxygen gas inside these snowlines (see Paper I for details). Radial drift and gas accretion move the N$_2$ snowline inward by up to $\sim$50\% compared to a static disk. The presence of N$_2$ in a water ice environment rather than as pure ice moves the N$_2$ snowline significantly inward by $\sim$70\%. Taken together, disk dynamics and ice morphology move the CO snowline inward by a factor of $\sim$7. The results of an enhanced gas-phase N/O ratio between the H$_2$O and CO snowlines compared to its average value, and of highly elevated N/O ratios in gas between the CO and N$_2$ snowlines (see Figure \ref{fig:Nstatic}), are preserved.}  
\label{fig:NO_ratio}
\end{figure}

@@

%\emgr{Discuss how entrapment of volatiles by H2O affects volatile abundances and C/O ratios. Re-emphasize the fact that the C/O and N/O ratios are upper estimates, and that CH4 and NH3 might matter in a viscous disk. State that we plan to address this in a future paper. More TBD.}
This study shows that the gas-phase N/O ratio in protoplanetary disks is considerably enhanced throughout most of the disk midplane compared to its average value. As demonstrated in Figure \ref{fig:enhance}, the gaseous N/O ratio is enhanced by a factor of two beyond the H$_2$O snowline, by more than a factor of three between the CO$_2$ and CO snowlines, and by several orders of magnitude between the CO and N$_2$ snowlines. Thus constraining the N/O ratio in a giant planet atmosphere could be used to trace its formation origins.

Theoretical models of the magnitude and role of N/O (and N/C) ratios in exoplanet atmospheres are therefore key in order to use these ratios as probes for a planet's formation location. Models that explore the effect of varying the C/O ratio in exoplanet atmospheres exist in literature, and they display a large and observable effect on gas giant envelope chemistry (\citealt{lodders09}, \citealt{molliere15}). However, no similar model explorations exist for the effect of N/O and C/N/O ratios, and both are needed to exploit this potential constraint. Given the existence of such theoretical models, measurements of the N/O ratio in planetary envelopes may be possible to infer from atmospheric compositions of nitrogen versus carbon and oxygen bearing species. Nitrogen carriers have not been targeted so far due to lack of instrument sensitivity, but such observations and detections are likely in the near future with the advent of JWST (e.g., NH$_3$, \citealt{greene16}). The N/O ratio enhancement is larger than that of the gas phase C/O ratio throughout most of the disk. Thus measurements of an enhanced C/O ratio in an exoplanet atmosphere could be corroborated (disproved) by measurements of enhanced (non-enhanced) N/O ratios. Moreover, Figure \ref{fig:enhance} shows that giant planets that have formed in situ between the H$_2$O and CO snowlines are expected to present elevated both C/O and N/O ratios in their atmospheres, whereas planets between the CO and N$_2$ snowlines will have a highly enhanced N/O ratio in their atmospheres, but not C/O.

@@

Due to disk dynamics and ice morphology, the disk regions with highly elevated gas phase N/O and C/O ratios are uncertain and may span tens of AU. Both ice morphologies discussed in this study, pure and water dominated ices, are plausible in protoplanetary disks and depend on whether H$_2$O and CO ices formed on similar timescales or successively (e.g., \citealt{garrod11}). Observations of protostellar cores show that a large fraction of CO is bound in a pure ice multilayer \citep{pontoppidan03}, but theoretical models also suggest an icy mantle structure where CO resides on a H$_2$O ice layer (e.g., \citealt{collings03}). One can also imagine a scenario where CO is in a water binding environment and N$_2$ is not. This could be attributed to the fact that H$_2$O may bind preferentially to CO than N$_2$, since both H$_2$O and CO are polar molecules while N$_2$ is not. It is also possible for N$_2$ ices to form later than CO (e.g., \citealt{pagani12}), and thus be deposited on the outer layers of the icy mantles which are typically water poor (e.g., \citealt{garrod11}). Detections of CO and N$_2$ snowlines in the same disk could potentially verify this scenario, since CO and N$_2$ desorb at substantially different temperatures, and therefore locations, if the former is water dominated and the latter is in a pure ice environment. The impact of the ice environment on the snowline location is much smaller in the case of CO$_2$ and NH$_3$, as their binding energies and behavior are closer to that of H$_2$O. No detailed measurements for the CH$_4$ binding energy in a water environment exist so far, but due to its low desorption temperature a similar behavior to that of CO and N$_2$ would be expected. While the presence of some carbon in the form of CH$_4$ only modestly affects our results, CH$_4$ may become important in disks where a large fraction of the CO abundance has been converted into hydrocarbons (e.g., \citealt{du15}).

Changes in stellar luminosity (e.g., \citealt{kennedy06}) and gas mass accretion rate (e.g., \citealt{chambers09}), as well as the evolution of icy dust particles due to grain growth and fragmentation (e.g., \citealt{birnstiel12}), may introduce additional uncertainties in the snowline locations, and thus the C/N/O ratios. Given the number of uncertainties in snowline locations, detections of snowlines in a sample of disks at different evolutionary stages are needed to provide observational constraints. Current detections include the H$_2$O snowline in TW Hya \citep{zhang13} and the CO snowline in TW Hya \citep{qi13} and HD 163296 \citep{qi15}. However, more snowline detections in disks of different ages are expected in future ALMA cycles.

 %The impact of the ice environment on the snowline location is much smaller in the case of CO$_2$ and NH$_3$, as their binding energies and behavior are closer to that of H$_2$O. No detailed measurements for the CH$_4$ binding energy in a water environment exist so far, but due to its low desorption temperature a similar behavior to that of CO and N$_2$ would be expected. While the presence of some carbon in the form of CH$_4$ only modestly affects our results, CH$_4$ may become important in disks where a large fraction of the CO abundance has been converted into hydrocarbons (e.g., \citealt{du15}).   %It is thus crucial to build a quantitative framework to explore the range of snowline locations and thus gaseous C/N/O ratios across disks with different dynamics and grain morphologies, as such models are currently lacking. %The theoretical expectations for N/O ratios in disks and giant planets may be further constrained by future JWST observations of exoplanet atmospheric spectra and detections of nitrogen bearing species (e.g., NH$_3$, \citealt{greene16}).

Uncertainties in snowline locations of this magnitude also affect interpretations of Solar system observations. Recent measurements of nitrogen abundance in comet 67P/Churyumov-Gerasimenko found a N$_2$/CO ratio $\sim 10^{-3}$ \citep{rubin15}. A low N$_2$/CO ratio is consistent with comets having formed inside the N$_2$ snowline where N$_2$ is still in the gas phase. Theoretical models suggest that Jupiter-family comets, such as 67P, originate from the Kuiper belt \citep{duncan97} (but see \citealt{rubin15} for alternative formation scenarios for 67P). It is thus possible, in principle, to use measurements of the N$_2$ abundance in Jupiter-family comets to determine where the N$_2$ snowline was located in our Solar system. However due to the uncertainty in the calculated location of the N$_2$ snowline (see Section \ref{sec:dynamic}), more detailed modeling is needed.

%Finally, detections of snowlines in disks that are reasonably well characterized could help further constrain the dynamics and ice morphology in the disks.

@@

\item The effect of CH$_4$ and NH$_3$ on the C/O and N/O ratios is small, even when we consider the maximum observed CH$_4$ and NH$_3$ abundances in protostellar cores. In this scenario, the gas phase C/O ratio increases by 8\% between the CO$_2$ and CH$_4$ snowlines, and  the gaseous N/O ratio increases by up to 18\% between the NH$_3$ and CO$_2$ snowlines. In both cases, large gas phase C/O and N/O ratios in the outer disk are preserved.
